---
title: "Assignment 2"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include = FALSE}
# library(ca)
# library(rgl)
library(reshape)
library(ggplot2)
# library(ggforce)

selikoff.dat <- readRDS("Data/selikoff.rds")
sibs.dat     <- readRDS("Data/siblings.rds")
expsibs.dat  <- readRDS("Data/expsibs.rds")
```

# Part 1: Dimension Reduction

## Question 1

## Question 2
### Setup
We are given:

$$
N = 
\begin{pmatrix}
4 & 8 & 8 \\
3 & 6 & -9
\end{pmatrix}
$$
Therefore, 
$$
N^T = 
\begin{pmatrix}
4 & 3 \\
8 & 6 \\
8 & -9
\end{pmatrix}
$$
### a)

#### i) Without R

$$
N^TN 
=
\begin{pmatrix}
4 & 3 \\
8 & 6 \\
8 & -9
\end{pmatrix}
\times
\begin{pmatrix}
4 & 8 & 8 \\
3 & 6 & -9
\end{pmatrix}
= \begin{pmatrix}
25 & 50 & 5 \\
50 & 100 & 10 \\
5 & 10 & 145
\end{pmatrix}
$$
The characteristic equation of $N^TN$ is:

$$
\begin{aligned}
-18000\lambda + 270\lambda^2 - \lambda^3 &= 0 \\
\implies -(\lambda - 150)(\lambda - 120) \lambda &= 0 \\
\end{aligned}
$$
which gives us $\lambda_1 = 150, \lambda_2 = 120, \lambda_3 = 0$.

#### ii) With R

```{r}
# Calculate the matrix
N <- matrix(c(4,3,8,6,8,-9), ncol = 3)
NT <- t(N)
NTN <- NT %*% N
# use eigen() to get the values and vectors
NTNevecs <- eigen(NTN)$vectors
NTNevals <- eigen(NTN)$values
# List the values and vectors together. Note: because the eigen() function automatically normalises
# the eigenvectors, I have multiplied the eigenvectors by the normalising constant that comes from 
# part i) in order to get matching results.
(NTNv1 <- list("Lambda1" = NTNevals[[1]], "EV1"= NTNevecs[,1]*sqrt(30)))
(NTNv2 <- list("Lambda2" = NTNevals[[2]], "EV2"= NTNevecs[,2]*sqrt(6)))
(NTNv3 <- list("Lambda3" = round(NTNevals[[3]]), "EV3"= round(NTNevecs[,3]*sqrt(5))))
```

### b)
#### i) Without R

$$
NN^T 
= 
\begin{pmatrix}
4 & 8 & 8 \\
3 & 6 & -9
\end{pmatrix}
\times
\begin{pmatrix}
4 & 3 \\
8 & 6 \\
8 & -9
\end{pmatrix} 
= 
\begin{pmatrix}
144 & -12 \\
-12 & 126
\end{pmatrix}
$$

The characteristic equation of $NN^T$ is:

$$
\begin{aligned}
18000 - 270\lambda + \lambda^2 &= 0 \\
\implies (\lambda - 150)(\lambda - 120) &= 0 \\
\end{aligned}
$$
which gives us $\lambda_1 = 150, \lambda_2 = 120$.

#### ii) With R
```{r}
N <- matrix(c(4,3,8,6,8,-9), ncols = 3)
NT <- matrix(c(4,8,8,3,6,-9), ncols = 2)
```

#### ii) With R

```{r}
N <- matrix(c(4,3,8,6,8,-9), ncol = 3)
NT <- t(N)
NNT <- N %*% NT
NNTevals <- eigen(NNT)$values
NNTevecs <- eigen(NNT)$vectors

(NNTv1 <- list("Lambda1" = NNTevals[[1]], "EV1"= NNTevecs[1,]))
(NNTv2 <- list("Lambda2" = NNTevals[[2]], "EV2"= NNTevecs[2,]))
```

### c)

We perform an _Singular Value Decomposition_ in __R__ by using the `svd()` function.
```{r}
(SVDN <- svd(N))
crossprod(NNTv1$EV1, SVDN$u[,1])
crossprod(NNTv2$EV2,SVDN$u[,2])
NNTv2$EV2%*%SVDN$u[,2]
```

# Part 2: Simple Correspondence Analysis

## Question 3

## Question 4

```{r, fig.width=4.5,fig.height=4.5, fig.align="centre"}
### Complex Elliptical
plotEllipse.exe <- function(df,             # an IxJ contingency table 
                            xlim = c(-1,1), # x limits for chart
                            ylim = c(-1,1),  # y limits for chart
                            CoordType = "Principal", # either Principal or Standard
                            ExtraAxes = TRUE # Take into account info from Principal Axes > 2
                            ){
  # Definitions:   
  N = sum(df) # Total observations
  P =  df/N   # Matrix of proportions
  M = min(nrow(df),ncol(df))-1 # The number of Principal Axes.
  dof = (nrow(df)-1)*(ncol(df) - 1) # degrees of freedom
  dI =  diag(apply(P, 1, sum),nrow(df) ,nrow(df)) # Diagonal matrix of P matrix row totals
  dJ =  diag(apply(P, 2, sum),ncol(df) ,ncol(df)) # Diag matrix of P matrix column totals
  rr =  rowSums(P) # Row sums of P matrix
  cc =  colSums(P) # Column sums of P matrix
  Z = solve(sqrt(dI))%*%(P - rr%*%t(cc))%*%solve(sqrt(dJ))
  svdZ = svd(Z) # Singular value decomposition of Z
  

  # Working:
  A = svdZ$u # row scores
  B = svdZ$v # col scores
  dimnames(A)[1] = dimnames(P)[1]
  dimnames(B)[1] = dimnames(P)[2]
  
  if(CoordType == "Standard"){
  F = svdZ$u
  G = svdZ$v
  dimnames(F)[1] = dimnames(P)[1]
  dimnames(G)[1] = dimnames(P)[2]
  ExtraAxes = FALSE # not sure about this. This edits the settings so the 
                    # ExtraAxes information isn't added when calculating
                    # the confidence region.
  }
  if(CoordType == "Principal"){
  dlambda = diag(svdZ$d)
  F = sqrt(solve(dI))%*%A%*%dlambda
  G = sqrt(solve(dJ))%*%B%*%dlambda
  dimnames(F)[1] = dimnames(P)[1]
  dimnames(G)[1] = dimnames(P)[2]
  }
  
  prininert = svdZ$d^2
   totinert = sum(svdZ$d^2)
       chi2 = totinert*N 
  princoord = as.data.frame(rbind(F[,1:2], G[,1:2]))
  
  if(ExtraAxes==TRUE){
  FF = vector(mode = "numeric", length = nrow(F))
  FFF = vector(mode = "numeric", length = M-3)
  extraF <- for(i in 1:nrow(F)){
    for(j in 3:M){
      FFF[j] = (F[i,j]/sqrt(prininert[j]))^2
      FF[i] = sum(FFF[j])
    }
    
  }
  GG = vector(mode = "numeric", length = nrow(G))
  GGG = vector(mode = "numeric", length = M - 3)
  extraG <- for(i in 1:nrow(G)){
    for(j in 3:M){
      GGG[j] = (G[i,j]/sqrt(prininert[j]))^2
      GG[i] = sum(GGG[j])
    }
    
  }
  } else {
    FF = rep(0,nrow(P))
    GG = rep(0, ncol(P))
  }
  
  sMaj = matrix(nrow = nrow(P), ncol = 2)
  for(i in 1:nrow(P)){
    sMaj[i,1] = sqrt(prininert[1])*sqrt((qchisq(0.95,dof)/chi2)*(1/rowSums(P)[i] - FF[i]))
    sMaj[i,2] = sqrt(prininert[2])*sqrt((qchisq(0.95,dof)/chi2)*(1/rowSums(P)[i] - FF[i]))
  }
  sMin = matrix(nrow = ncol(P), ncol = 2)
  for(i in 1:ncol(P)){
    sMin[i,1] = sqrt(prininert[1])*sqrt((qchisq(0.95,dof)/chi2)*(1/colSums(P)[i] - GG[i]))
    sMin[i,2] = sqrt(prininert[2])*sqrt((qchisq(0.95,dof)/chi2)*(1/colSums(P)[i] - GG[i]))
  }
  
    sMajMin = rbind(sMaj,sMin)
  chartdata = cbind(princoord,sMajMin)
  
  ellipse <- function(sMaj,sMin, xc, yc, res = 100){
    xp <- NULL
    yp <- NULL
    for(i in 0:res){
      a  = (2 * pi * i)/res
      x  = sMaj * cos(a)
      y  = sMin * sin(a)
      xp = c(xp, x + xc)
      yp = c(yp, y + yc)
    }
    lines(xp, yp, type = "l", col = "light green", lwd = 0.6)
    invisible()
  }
  plot(xlim, ylim, 
       xlab = paste("First Principal Axis:", round(prininert[1]*100/totinert,2),"%"),
       ylab = paste("Second Principal Axis:", round(prininert[2]*100/totinert,2),"%"),
       type="n")
  title(main = "Elliptical Confidence Intervals",
        sub = list(paste("% Association attributed to Third Principal Axis and higher:",
                         round(sum(prininert[3:length(prininert)])*100/totinert,2),
                         "%"),
                   cex = 0.8))
    abline(v = 0, col = "grey")
    abline(h = 0, col = "grey")
    points(chartdata[,1:2], col = "red")
  for(i in 1:nrow(chartdata)){
    ellipse(sMaj = chartdata[i,3],
            sMin = chartdata[i,4],
            xc = chartdata[i,1],
            yc = chartdata[i,2])
  }
  text(chartdata[,1:2], labels=rownames(chartdata), adj = c(0.5,-1),cex= 0.8)
  text(xlim[1], ylim[2], labels = paste("Coordinate Type:",CoordType), adj = c(0.05,0.5), cex = 0.7, col = "dark grey")
}

plotEllipse.exe(selikoff.dat, xlim = c(-1,2))
plotEllipse.exe(selikoff.dat, ExtraAxes = FALSE, xlim = c(-1,2))
plotEllipse.exe(selikoff.dat, CoordType = "Standard")
```




# Part 3: Multiple Categorical Variables

### e)
Write an R function that performs a simple correspondence analysis on an IxJ contingency table. Make sure that the output consists of the principal coordinates, the inertia values and a two-dimensional correspondence plot.

```{r}
## TURN INTO FUNCTION
# Selikoff's Data

svdZ = svd(Z)
A = svdZ$u
dimnames(A)[1] = dimnames(P)[1]
B = svdZ$v
dimnames(B)[1] = dimnames(P)[2]

# Principal Inertia
svdZ$d^2
# Total Inertia
sum(svdZ$d^2)

# Percentage contribution to the total inertia of each axis
100*svdZ$d^2/sum(svdZ$d^2)

# Cumulative % contribution to the total inertia of each axis
cumsum(100*svdZ$d^2/sum(svdZ$d^2))

# Principal Coordinates

dlambda = diag(svdZ$d)
F = sqrt(solve(dI))%*%A%*%dlambda
dimnames(F)[1] = dimnames(P)[1]
F
G = sqrt(solve(dJ))%*%B%*%dlambda
dimnames(G)[1] = dimnames(P)[2]
G
```


## Question 5
```{r}
sib <- c(15,31,35,18,34,60,45,14,36,46,30,3,22,25,13,3,61,26,8,4,
         17,60,63,15,53,96,74,15,70,45,39,9,67,40,24,2,79,31,7,1,
         7,5,5,2,20,12,10,1,23,11,4,2,16,12,4,0,36,7,3,1)
sib <- array(sib, dim = c(4,5,3))

dimnames(sib) <- list("Years Of Schooling"=c("<12","12","13-16","17+"), "Number of Siblings"=c("0-1","2-3","4-5","6-7","8+"), Happiness=c("Not Too Happy","Pretty Happy","Very Happy"))
sib
```

```{r Q5Func, warning = FALSE}
# DONE
ChisqFunc3D.exe <- function(array3){
  
  options(warn = FALSE)
  
  XIJ  = apply(array3, c(1,2), sum)
  XIK  = apply(array3, c(1,3), sum)
  XJK  = apply(array3, c(2,3), sum) 
  XIJK = apply(array3,2,rbind)
  
  chisqIJ  = chisq.test(XIJ)
  chisqIK  = chisq.test(XIK)
  chisqJK  = chisq.test(XJK)
  chisqIJK = chisq.test(XIJK)
  
  statIJ  = chisqIJ$statistic
  statIK  = chisqIK$statistic
  statJK  = chisqJK$statistic
  statIJK = chisqIJK$statistic
  statSum = sum(statIJ, statIK, statJK, statIJK)
  
  dfIJ  = chisqIJ$parameter
  dfIK  = chisqIK$parameter
  dfJK  = chisqJK$parameter
  dfIJK = chisqIJK$parameter
  dfX2  = prod(dim(array3)-1)
  
  pIJ  = round(chisqIJ$p.value,3)
  pIK  = round(chisqIK$p.value,3)
  pJK  = round(chisqJK$p.value,3)
  pIJK = round(chisqIJK$p.value,3)
  pX2  = round(1-pchisq(statSum, dfX2),5)
  
  out   = array(c(statIJ,statIK,statJK,statIJK,statSum,
                 dfIJ,dfIK,dfJK,dfIJK,dfX2,
                 pIJ,pIK,pJK,pIJK,pX2),
               dim = c(5,3))
  
  dimnames(out) = list(c("XIJ","XIK","XJK","XIJK","X2"),
                      c("X2","DoF","P.value"))
out
}

(ChisqFunc3D.exe(sib))
```


## Question 6

```{r, eval = FALSE, echo = TRUE}
# mjca(expsibs.dat)
# names(mjca(expsibs.dat))
mjca_expsibs <- mjca(expsibs.dat)
```

