---
title: "R Notebook"
output: html_notebook
---

## Pearson’s Statistic for a 2x2 Table

### Asbestos Example

```{r}
options(digits = 9)
asbestos.dat <- matrix(c(522,53,203,339), nrow = 2)
dimnames(asbestos.dat) = list(c("0-19","20+"),c("No","Yes"))
asbestos.dat

# matrix of proportions (matrix / sample size)
asbestos.dat/1117

#apply(asbestos.dat/1117,1,sum)
rowSums(asbestos.dat)/1117

#apply(asbestos.dat/1117,2,sum)
colSums(asbestos.dat)/1117

# Chi-Squared Statistic computation for the asbestos.dat data.
1117*(0.46732319-0.64906*0.5147717)^2/(0.64906 + *0.35094*0.5147717*0.4852283)

chisq.test(asbestos.dat, correct = FALSE)
```

Here, correct = F has been specified to ensure Yates’ continuity correction (see next slides) is not imposed – by default the correction is incorporated into the analysis of a 2x2 contingency table.

Yates (1934) argued that the chi-squared statistic with 1 degree of freedom gives p-values from Pearson’s statistic that typically underestimate the true p-values. As a result he proposed an adjustment to Pearson’s chi-squared statistic of a 2x2 contingency table (see lecture notes).

```{r}
# Including Yates' continuity correction.
chisq.test(asbestos.dat)

# Calcluation including Yates continuity correction.
1117*(abs(0.46732319-0.64906*0.5147717) - 0.5/1117)^2/(0.64906*0.35094*0.5147717*0.4852283)
```


## Approximations of the Tetrachoric Correlation

To overcome the computational difficulties that he faced at the time, Pearson (1900, pg 7) proposed a number of approximations to his correlation (see notes).

```{r}
library(psych) # required for the tetrachoric function
tetrachoric(asbestos.dat)
```

```{r}
selikoff.dat <- matrix(c(310,212,21,25,7,36,158,35,102,35,0,9,17,49,51,0,0,4,18,28), nrow = 5)
dimnames(selikoff.dat) <- list(c("0-9","10-19","20-29","30-39","40+"),c("None","Grade 1","Grade 2","Grade 3"))
selikoff.dat
rowSums(selikoff.dat)
colSums(selikoff.dat)
```


```{r}
oddsratio.exe <- function (N, acc) {
  round((N[1,1]*N[2,2])/(N[1,2]*N[2,1]), digits = acc)
}
oddsratio.exe(asbestos.dat, acc = 3)

edwards.odds.exe <- function(N, b, acc){
  OR = oddsratio.exe(N, acc)
  EO = round((OR^b - 1)/(OR^b+1), digits = acc)
  VarO = ((4*b^2*OR^(2*b))/(OR^b + 1)^4)*(N[1,1] + N[1,2] + N[2,1] + N[2,2])^(-1)
  return(list("OR Value" = EO, "OR Variance" = VarO))
}

# Yule's Q
edwards.odds.exe(asbestos.dat, b = 1, acc = 4)

# Yule's Y
edwards.odds.exe(asbestos.dat, b = 0.5, acc = 4)

# Digby's H
edwards.odds.exe(asbestos.dat, b = 3/4, acc = 4)

# Edward's J
edwards.odds.exe(asbestos.dat, b = pi/4, acc = 4)
```

## Breslow-Day Test

```{r}
breslowday.test = function(x, OR=NA){
 ## function to compute the Breslow Day test of homogeneity
 ## for a 2 by 2 by k table
 ## x is a three dim array, 2x2xk
 if(is.na(OR)) {
   pooled = apply(x,1:2,sum)
   OR = pooled[1,1] * pooled[2,2] / pooled[1,2] / pooled[2,1]
 }
 k = dim(x)[3]
 n11k = x[1,1,]
 n21k = x[2,1,]
 n12k = x[1,2,]
 n22k = x[2,2,]
 row1sums = n11k + n12k
 row2sums = n21k + n22k
 col1sums = n11k + n21k
 Amax = apply(cbind(row1sums,col1sums),1,min)
  ## Astar must be no more than col1sums and no more than row1sums
 bb = row2sums +row1sums * OR - col1sums*(1-OR)
 determ = sqrt(bb^2 + 4 * (1-OR) *  OR * row1sums * col1sums)
 Astar = (-bb + cbind( -determ, determ))/ (2 -2*OR)
 Astar = ifelse(Astar[,1] <= Amax & Astar[,1] >= 0, Astar[,1], Astar[,2])
 # print(Astar)
 Bstar = row1sums - Astar
 Cstar = col1sums - Astar
 Dstar = row2sums - col1sums + Astar
 Var = apply(1 / cbind(Astar,Bstar,Cstar,Dstar), 1, sum)^(-1)
 # print(Var)
 X2 = sum( (x[1,1,] - Astar)^2/Var )
 pvalue = 1 - pchisq(X2,k-1)
 return(unlist(list(OR = OR, Stat = X2, df = k-1, pvalue = pvalue)))
}

biomassTB.Salary = array(c(38,12,102,141,12,9,136,383),
                         dim=c(2,2,2),
                         dimnames=list(exposure=c("exposed","not"),
                                       disease=c("case","control"),
                                       salary=c("<1000",">=1000")))

breslowday.test( biomassTB.Salary)
#  breslowday.test( biomassTB.Salary, 4.02)
#        OR       Stat         df     pvalue 
#4.15000000 0.07150466 1.00000000 0.78915846 

```

